---
title: "{{< var course.code >}}: Week 1 Lab — R from Zero to Real Data"
subtitle: "{{< var course.institution >}} · {{< var course.semester >}}"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

# Before lab: install R and RStudio

Do this before you arrive if you can; it takes about ten minutes.

1.  **Install R** — the language itself — from CRAN:
    <https://cran.r-project.org>. Choose your operating system and
    install the latest release with the default options.
2.  **Install RStudio Desktop** — the environment we write R in — from
    Posit: <https://posit.co/download/rstudio-desktop/>.

Order matters: R first, then RStudio. RStudio is not R; it is a workshop
built around R. If you open RStudio and see a console with a `>` prompt,
both installations worked.

# A quick tour of RStudio

Open RStudio. You will see up to four panes:

- **Console** (left or bottom-left): where R actually runs. Type a
  command at the `>` prompt, press Enter, get an answer.
- **Source** (top-left, appears when you open a file): where scripts
  live — files of commands you keep.
- **Environment** (top-right): the objects you have created this
  session.
- **Files / Plots / Help** (bottom-right): your file system, your
  figures, and the documentation.

## The console: R as a calculator

Click in the console and try a few lines. R evaluates each and prints
the result:

```{r calculator}


```

The `[1]` at the front of each answer is R telling you the position of
the first printed value — it matters when results have many values, as
they soon will.

You store results with the assignment arrow `<-` ("gets"):

```{r assignment}



```

Three things just happened that you will do thousands of times this
semester: you **combined** values into a vector with `c()`, you
**named** the result with `<-`, and you **applied a function** —
`mean()` — to it. Everything in R is a variation on this pattern.

## Scripts: why we never work in the console alone

Anything you type in the console vanishes when you close RStudio. So we
work in **scripts**: *File → New File → R Script*. Type commands in the
script, put your cursor on a line, and press **Ctrl+Enter** (Mac:
**Cmd+Return**) to send it to the console. Save the script with a name
like `lab01.R`.

::: callout-note
## Why scripts are the whole game

The script *is* your analysis. If your script runs top to bottom on a
fresh session and produces your results, your work is reproducible: you
can rerun it, your co-author can rerun it, and reviewers can rerun it.
If your results depend on things you clicked or typed once and forgot,
they are not results — they are anecdotes. Every empirical claim you
make in this course should trace back to a script.
:::

Two console habits to build immediately: the **Up arrow** recalls
previous commands, and `?mean` (a question mark before any function
name) opens its help page.

# Packages: installing the tidyverse

Base R can do everything we need, but the **tidyverse** — a suite of
packages with a consistent grammar for data work — is the dialect this
course speaks. Install it once per machine:

```{r install}
#| eval: false
install.packages("tidyverse")   # once per computer; note the quotes
```

Give it a few minutes. Then, in *every* R session where you want to use
it:

```{r library}
library(tidyverse)   # once per session; no quotes needed
```

The distinction trips everyone up at first: `install.packages()` buys
the book (once); `library()` takes it off the shelf (every session).
When you see the error `could not find function "read_csv"`, the
diagnosis is almost always a missing `library(tidyverse)` at the top of
your script.

When you load the tidyverse you will see a burst of startup messages
listing the attached packages and a couple of "conflicts." That is
normal, not an error.

# Projects and relative paths

Before touching the data, we set up the habit that prevents 90% of
beginner misery: the **RStudio Project**.

A project is just a folder that RStudio treats as home base. Create one
for this course: *File → New Project → New Directory* (or *Existing
Directory* if you already made a course folder). Inside it, mirror the
course layout:

```         
gov6019/                  <- your project lives here
  data/                   <- voter2012.csv and its codebook
  scripts/                <- gov6019_ggplot.R (the course figure theme)
  weeks/
    w01_intro_data/       <- this document, your lab01.R script
```

Everything in `data/` and `scripts/` downloads from the course site —
grab `voter2012.csv`, its codebook, and `gov6019_ggplot.R` now. (The
script becomes essential in Week 2, when we start making figures.)

Why bother? Because of **paths**. Every file on your computer has an
absolute path (`C:/Users/you/Desktop/gov6019/data/voter2012.csv` or
`/Users/you/gov6019/data/voter2012.csv`) — and absolute paths are poison
in a script, because they work on exactly one computer: yours, today. A
**relative path** (`data/voter2012.csv`) works for anyone who has the
folder.

::: callout-important
## Setting working directory

One subtlety worth learning now, because it will otherwise bite you at a
worse time: **where a relative path starts from depends on where the
code runs.** Type `getwd()` in the console to see your current working
directory. If it doesn't match your project root, you can type 
`setwd("path/to/project")`

```{r}

getwd()

```

I am already in the right folder. If I wasn't, I would set the working directory 
by locating where the file is on my computer.

```{r}

setwd("/Users/izzamalik/Desktop/cornell/govt_6019")

```



# Loading the VOTER data

The course's running dataset is `voter2012.csv`: a subset of the **2012
VOTER Study**, a large national opinion survey of U.S. adults (n =
8,000). We will live with this dataset for most of the semester — by
Week 8 you will know its every quirk, and it has quirks.

Load it with `read_csv()`:

```{r load}
#| message: TRUE
voter <- read_csv("voter2012.csv")
```

::: callout-tip
## Which path goes in *your* script?

Typing along in your own `lab01.R`, run from your project's console?
Then the path is `data/voter2012.csv` — the `../../` form above is what
*this document* uses, because it starts from its own folder two levels
down. This is the console-versus-document distinction from the previous
section, live.
:::

The message is worth reading, not skipping: 8,000 rows, 9 columns, and
every column parsed as `dbl` (a number). Keep that last fact in mind.

## First looks

Four inspection commands you will run on every dataset you ever meet:

```{r inspect-dim}
 # rows, columns

 # variable names
```

```{r inspect-glimpse}
  # every variable, its type, and its first values

```

```{r inspect-head}
 # the first six rows, as a table
```

Look at the actual rows. Every value is a bare integer. Is a `pid3` of
`1` a Democrat or a Republican? Is `gender` `2` a woman or a man? Does
`important_econ = 1` mean the economy is very important or not at all?
**The data will not tell you.** This is how survey data actually
arrives: the numbers live in the file and the meaning lives somewhere
else.

## The codebook-first habit

The meaning lives in the **codebook**. Make sure to review it first before 
getting started on working with the data. Here is its variable table:

| Variable | Question / meaning | Values |
|------------------------|------------------------|------------------------|
| `important_econ` | "The economy is an important issue" (agreement) | 1 Strongly agree · 2 Agree · 3 Disagree · 4 Strongly disagree · 5 Don't know |
| `important_gayr` | "Gay rights are an important issue" (agreement) | same scale |
| `unimportant_budget` | "The budget deficit **is not** an important issue" (agreement) | same scale |
| `unimportant_abort` | "Abortion **is not** an important issue" (agreement) | same scale |
| `educ` | Highest education completed | 1 No HS · 2 HS grad · 3 Some college · 4 2-year · 5 4-year · 6 Post-grad |
| `birthyr` | Year of birth | 1921–1994 |
| `pol_interest` | Interest in politics and current affairs | 1 Not much · 2 Somewhat · 3 Very much · 4 Not sure |
| `pid3` | Party identification | 1 Democrat · 2 Republican · 3 Other |
| `gender` | Gender | 1 Man · 2 Woman |

Empty cells in the CSV are item nonresponse — people who skipped the
question — and R reads them as `NA`, its symbol for "missing."

::: callout-important
## Rule: codebook before computation

Never compute anything on a variable you have not looked up in the
codebook. In this dataset the columns are numbers, but — think back to
lecture — most of these variables are not *interval* measurements.
`pid3` is nominal: its numbers are arbitrary labels. `educ` and the
issue items are ordinal: order means something, distances do not. Only
`birthyr` is a genuine number in the arithmetic sense. The codebook is
what tells you which is which.
:::

# Five verbs: your first real wrangling

The tidyverse organizes data manipulation around a small set of verbs,
each of which takes a dataset and returns a dataset. We will chain them
with the **pipe**, written `|>`, which you can pronounce as "then":

```{r pipe-intro}
 # exactly the same as dim(voter): take voter, THEN apply dim

```

The pipe earns its keep when steps stack up, which is most of the time.

## `select()`: choose columns

```{r select}

```

## `filter()`: choose rows

`filter()` keeps rows meeting a logical condition. Note the double
equals `==` for "is equal to"

```{r filter}
# respondents who report being very interested in politics (3 from codebook)
# double equals sign (==) means is equal to


```

5,063 of the 8,000 respondents. Conditions can be combined with commas
(meaning "and"):

```{r filter-and}
# very interested in politics AND identifying as democrats (both conditions must be true)



```

## `arrange()`: sort rows

```{r arrange}
# youngest respondents first; sort by birth year, descending

```

Drop the `desc()` and you sort ascending — you will need that in the
exercises.

## `mutate()`: create new variables

Here is the first of the codebook's little tests. There is no age
variable — only `birthyr`. But the survey was fielded in 2012, so age is
one subtraction away:

```{r mutate}

```

Two things to notice. First, we assigned the result back to `voter` —
without `voter <-` at the front, the new column is computed, printed,
and thrown away. Second, this age is a good approximation, not an exact
one; someone born in late 1994 who had not yet had their 2012 birthday
is 17, not 18. For describing a dataset that is fine; the honest move is
to know it and say it.

```{r age-range}
# $ pulls out a single column 


```

Ages 18 to 91 — consistent with a survey of adults, which is exactly the
kind of sanity check you should run on every variable you derive.

## `count()`: tabulate

The workhorse for categorical variables:

```{r count-pid3}

```

Read this *with the codebook*: 3,639 Democrats, 3,091 Republicans, 1,181
identifying with neither major party — and 89 people with `NA`, who
skipped the question. `count()` shows missing values as their own row
rather than hiding them, which is the correct instinct and one we will
keep.

```{r count-educ}

```

Again: this table is unreadable without the codebook, and perfectly
clear with it (`6` = post-grads: 1,210 of them, which will matter in the
exercises).

# A number that lies

Time to compute our first substantive summary. The economy item asks for
agreement with "The economy is an important issue." What was the average
response in 2012, three years after the Great Recession?

```{r mean-na}

```

`NA`. R is not broken; it is being scrupulously honest. Ninety-seven
respondents skipped this question, so their values are missing — and R
will not average numbers it does not have unless you explicitly tell it
how to handle the gaps. 

```{r number of missing values}


```


The standard instruction is `na.rm = TRUE`
("remove `NA`s before computing"):



```{r mean-narm}


```

There we go: **1.14**, on a scale where 1 = *Strongly agree*. In 2012,
with unemployment near 8%, respondents all but unanimously agreed the
economy was= an important issue. Plausible. Publishable, even.

Before we congratulate ourselves, one more habit: never summarize a
variable you have not tabulated. What values does `important_econ`
actually take?

```{r count-econ}

```

Look at the fifth row. Thirty-nine people have the value **5** — on what
the codebook describes as a four-point agreement scale. The economy:
apparently so important that 39 respondents rated it a five out of four.

Scroll back to the codebook table. The fifth value is right there: **5 =
Don't know.**

Now think about what our mean just did. It treated "Don't know" as a
*substantive* answer — one notch beyond *Strongly disagree*, as if
declining to answer were the most extreme disagreement on the scale.
That is nonsense: "Don't know" is nonresponse wearing a number. It
belongs with the `NA`s, but because it arrived coded as `5`,
`na.rm = TRUE` sailed right past it. Our 1.14 is contaminated — only
mildly here, since the 5s are 39 answers out of nearly 8,000, but wrong
in a way that **no error message will ever flag**. `NA` was missingness
R could see. The 5s are missingness in disguise.

And it is not just this variable:

```{r count-others}

```

`pol_interest` runs 1–3 ("Not much" to "Very much") — and then 57 people
at `4`, which the codebook reveals to be *Not sure*. Same trap,
different costume. The other three issue items carry the 5 = Don't know
code too, and two of them (`unimportant_budget`, `unimportant_abort`)
have a further problem you may have noticed in their wording that we
have not even touched yet.

::: callout-warning
## Until further notice

Do **not** report a mean, or any other summary, of the four issue items
or of `pol_interest` from the raw data. Every such number is
contaminated by disguised nonresponse, and the fix — telling R which
codes are really missing, then dealing with what the recoding reveals —
is precisely next week's lab. Until then, treat those five columns as
radioactive: you may `count()` them, and nothing else.

To be continued.
:::

# Exercises

Now you. Each exercise gives you scaffolded code with blanks (`___`);
replace the blanks and run it. The chunks are marked not to run as-is —
they will error until you fill them in. Fight with each one for a few
minutes before opening the solution.

## Party identification by gender

How does party identification break down by gender? Produce one table
with a row for every combination. (Hint: `count()` accepts more than one
variable.)

```{r ex1}
#| eval: false
voter |>
  count(___, ___)
```

In words, with the codebook at hand: which party's identifiers include
more women than men?


## Post-graduates

Two counts: (a) how many respondents hold a post-graduate degree? And
(b) how many of them are women? Check the codebook for the right codes
before you type.

```{r ex2}
#| eval: false
# (a)
voter |>
  filter(___) |>
  dim()

# (b)
voter |>
  filter(___, ___) |>
  dim()
```



## The oldest respondent

Who is the oldest respondent in the data — birth year, age in 2012,
gender, and party? Sort the data so the oldest respondents come first
and inspect the top of the result. (Hint: oldest means the *smallest*
birth year.)

```{r ex3}
#| eval: false
voter |>
  arrange(___) |>
  select(birthyr, age, gender, pid3) |>
  head(3)
```



## Age in decades

Create a variable `age_decade` giving the *decade of age* each
respondent is in: 20 for ages 20–29, 30 for ages 30–39, and so on. Then
tabulate it. The tool: integer division `%/%`, which divides and drops
the remainder (`47 %/% 10` is `4`).

```{r ex4}
#| eval: false
voter |>
  mutate(age_decade = ___) |>
  count(___)
```



# Checklist

By the end of this lab you should be able to:

- [ ] Open RStudio, write commands in a script, and run them with
  Ctrl+Enter / Cmd+Return — and say why scripts beat the console.
- [ ] Explain the difference between `install.packages()` (once per
  machine) and `library()` (every session).
- [ ] Work inside an RStudio Project and load files with relative paths
  — and state the course rule about `setwd()` and absolute paths.
- [ ] Load a CSV with `read_csv()` and inspect it with `dim()`,
  `names()`, `glimpse()`, and `head()`.
- [ ] Explain why a dataset of bare integers is meaningless without its
  codebook, and make checking the codebook your reflex.
- [ ] Use `select()`, `filter()`, `arrange()`, `mutate()`, and
  `count()`, chained with `|>`.
- [ ] Derive a new variable (like `age`) with `mutate()` and
  sanity-check it.
- [ ] Explain what `NA` is, what `na.rm = TRUE` does — and why it was
  *not* enough to make `mean(important_econ)` trustworthy.

That last box is the one to sit with. You watched a clean-looking,
plausible-sounding number get quietly poisoned by a code the software
had no way to recognize as missing. Next week we do something about it.
