---
title: "{{< var course.code >}}: Homework 4 — Expectation"
subtitle: "{{< var course.semester >}}"
author:
  - "{{< var course.instructor >}} (Instructor)"
  - "{{< var course.ta >}} (Teaching Assistant)"
format:
  pdf:
    documentclass: article
    pdf-engine: pdflatex
    fontsize: 11pt
    geometry: margin=1in
    colorlinks: true
    include-in-header:
      - file: ../../theme/macros.tex
execute:
  echo: true
  eval: false
---

<!-- IN PROGRESS. The set is being built up as Week 4 is drafted; more
questions to come. Five placed as of 2026-09-25.

Q4 "Unseen vignettes" was added 2026-09-22 (user instruction). It practises
indicator variables, P(A) = E[I_A], and linearity without independence. It is
the birthday problem with the days replaced by vignette versions.

Q1 "Name the distribution" is inherited verbatim from hw_w03 Q6, moved here on
2026-09-19. It tests the W3 discrete zoo and sets up the "From Bernoulli
to binomial" frame in the W4 deck. It opens the set as a review item on user
instruction, 2026-09-21.

Q2 must NOT name Simpson's paradox anywhere in the prompt. Part (g) is the
whole point of the question and is spoiled by naming it earlier. See
_HELD-FOR-MIDTERM.md and _NOTES-FOR-2027.md.

Q2(e) is written in sample notation only. E[Y|X] and the law of iterated
expectations left the W4 deck for Week 5 on 2026-09-21, so students have not
met that notation when this set is due. Do not "improve" it back.

Q3(g) was added 2026-09-25 (user instruction) when the "Shifting and
rescaling" frame was cut from the deck. Students derive E[a + bU] = a + bE[U]
from the three properties; the old (g) became (h) and uses that result.

Q5 "Contact and turnout" was added 2026-09-25 (user instruction) as Part E.
It covers the Variance and covariance section of the deck. The normal
distribution, pnorm and qnorm are held for Homework 5, after the CLT.

Solutions: hw_w04_solutions.qmd, which must be kept in step with this file. -->

# Part A — Review: naming a distribution

## Question 1: Name the distribution

For each scenario, name the distribution the random variable follows and give
its parameters, with one sentence of justification.

i. A state statute requires that the order of the 7 candidates on the ballot
   be randomized, with every ordering equally likely. $X$ is the ballot
   position assigned to one named candidate.

ii. One voter is sampled at random from a district. $X = 1$ if they support
    the incumbent and $X = 0$ otherwise.

iii. Fifteen voters are sampled independently at random from a district in
     which 40 percent of voters support the incumbent. $X$ is the number of
     sampled voters who support the incumbent.

iv. Motions of no confidence are tabled in a parliament at an average rate of
    1.5 per year. $X$ is the number tabled in a given year.

Then compute the following.

a. For (i): the probability that the named candidate is assigned the top
   position, and the probability that they are assigned one of the top two.

b. For (iii) with four sampled voters rather than fifteen: list every sequence
   of four responses containing exactly two supporters, count them, and
   confirm your count against $\binom{4}{2}$. Then compute
   $\Prob(X = 2)$ by hand.

c. For (iii) as stated, with fifteen sampled voters: compute $\Prob(X = 6)$
   and $\Prob(X \le 5)$ in R.

   ```{r}
   # dbinom() gives the PMF; pbinom() gives the CDF
   ```

# Part B — Conditioning and pooled comparisons

## Question 2: Two ways to reach the hospital

A regional trauma registry records every patient transported to hospital after
a road accident, the mode of transport, the severity of the accident, and
whether the patient died. Accidents are classified as serious or less serious
at the scene, before transport is chosen. (The counts are stipulated for the
exercise.)

\begin{center}
\begin{tabular}{@{}lrrr@{}}
\toprule
Deaths / patients & Serious & Less serious & \textbf{All} \\
\midrule
Helicopter & 48/100 & 16/100 & \textbf{64/200} \\
Ambulance  & 60/100 & 200/1000 & \textbf{260/1100} \\
\bottomrule
\end{tabular}
\end{center}

Let $Y = 1$ if a patient died and $Y = 0$ otherwise. Let $X$ denote severity,
taking the values serious and less serious.

a. Compute $\bar Y$ for each mode of transport within each severity group.
   State which mode has the lower death rate among serious accidents, and
   which has the lower death rate among less serious accidents.

b. Compute $\bar Y$ for each mode of transport across all its patients. State
   which mode has the lower death rate overall.

c. Your answers to (a) and (b) disagree. Before explaining why, write each of
   the two pooled rates in (b) as a weighted average of that mode's two group
   rates from (a). Give the four weights as fractions, and say what each
   weight is the proportion of.

d. The two modes have different weights. Name the feature of the transport
   system that produces that difference, using the sentence in the problem
   about when severity is assessed.

e. Lecture wrote a pooled mean as
   $\bar Y = \sum_x \bar Y_{X=x}\,\hat p(x)$. Using that expression, state the
   two things that can differ between two groups of patients with different
   pooled rates. Say which one differs here.

f. A newspaper reports the two numbers from (b) under the headline
   "Helicopter transport saves lives." State the question that comparison
   actually answers, and the question a reader takes it to answer.

g. This problem demonstrates a paradox that was discussed in Week
   {{< var wk.data >}}. Name it. The Week {{< var wk.data >}} example concerned
   income and vote choice measured on individuals and on states. State what
   that example and this one have in common, in terms of $X$, $Y$, and the
   weights in (c).

# Part C — Expected utility

## Question 3: The utility of voting

Lecture put a decision model to a pedestrian choosing between jaywalking and
waiting. Put the same model to a citizen choosing between voting and
abstaining on election day.

A citizen prefers one of the two candidates. Voting costs them $C = 5$, paid
with certainty: the travel, the queue, the time off work. The vote changes the
outcome of the election only if it is decisive, which occurs with probability
$p$. If the vote is decisive, the citizen's preferred candidate takes office
instead of the other one, which they value at $B = 500$. Abstaining costs
nothing and changes nothing. Take $p = 1/50{,}000$.

a. Let $U_{\text{vote}}$ be the citizen's utility from voting. Write it as a
   random variable: give the two values it takes and the probability of each.
   Do the same for $U_{\text{abstain}}$.

b. Compute $\E[U_{\text{vote}}]$ and $\E[U_{\text{abstain}}]$. State which
   action the model predicts.

c. Rewrite $\E[U_{\text{vote}}]$ and $\E[U_{\text{abstain}}]$ in the
   parameters $p$, $B$ and $C$. Give the condition on those three quantities
   under which the model predicts voting.

d. Solve that condition with equality for $p$. Report the value of $p$ at
   which the citizen is indifferent, given the $B$ and $C$ above. One vote in
   a national electorate is decisive with probability on the order of
   $10^{-7}$. State what the model predicts for such an electorate.

e. The model in (c) predicts near-universal abstention, and turnout in
   national elections is not near zero. One repair adds a term $D$, the
   utility of the act of voting itself, received whether or not the vote is
   decisive:
   $$
   \E[U_{\text{vote}}] = pB - C + D .
   $$
   Give the condition on $D$ under which the model predicts voting, and
   compute the threshold value of $D$ using the numbers above.

f. Each of $p$, $B$, $C$ and $D$ differs across citizens. State in one
   sentence what a model of this form predicts about turnout across a
   population.

g. Lecture gave three properties of expectation: $\E[c] = c$;
   $\E[cX] = c\,\E[X]$; and $\E[X + Y] = \E[X] + \E[Y]$. Using only these,
   show that $\E[a + bU] = a + b\,\E[U]$ for any constants $a$ and $b$. Write
   one step per line. Name the property each step uses.

h. Suppose every citizen's utility is measured on a new scale,
   $U' = a + bU$ with $b > 0$, applied to both actions. Using the result in
   (g), show that the action the model predicts in (c)
   is unchanged. Then state why adding a constant to the utility of voting
   alone is a different operation, and name the term in (e) that performs it.

# Part D — Indicator variables

## Question 4: Unseen vignettes

A survey experiment has $k = 10$ versions of a vignette. Each of $n = 20$
respondents is assigned one version, independently and uniformly at random.

a. For version $j$, define $I_j = 1$ if no respondent sees version $j$, and
   $I_j = 0$ otherwise. Find $\Prob(I_j = 1)$ and $\E[I_j]$.

b. Let $U$ be the number of versions that no respondent sees. Write $U$ as a
   sum of the indicators in (a), and find $\E[U]$.

c. Are $I_1$ and $I_2$ independent? Explain why your answer to (b) does not
   depend on the answer.

d. For each pair of respondents, define an indicator equal to 1 if the two
   saw the same version. Find the expected number of pairs who saw the same
   version.

e. Simulate the assignment 10,000 times in R. Compare the mean of $U$ with
   your answer to (b). Report the share of simulated experiments in which at
   least one version goes unseen.

   ```{r}
   # sample(1:10, 20, replace = TRUE) assigns 20 respondents to 10 versions
   # replicate() repeats an expression; unique() lists the versions that appear
   ```

# Part E — Variance and covariance

## Question 5: Contact and turnout

A campaign's field team contacts some registered voters before an election.
For a registered voter drawn at random, let $C = 1$ if the voter was contacted
and $C = 0$ otherwise. Let $T = 1$ if the voter turned out and $T = 0$
otherwise. The joint distribution is below. (The probabilities are stipulated
for the exercise.)

\begin{center}
\begin{tabular}{@{}lrr@{}}
\toprule
 & $T = 0$ & $T = 1$ \\
\midrule
$C = 0$ & .30 & .40 \\
$C = 1$ & .05 & .25 \\
\bottomrule
\end{tabular}
\end{center}

a. Find $\E[C]$. Find $\V[C]$ from the definition
   $\V[X] = \sum_i p_i\,(x_i - \mu)^2$. Check your answer against the
   Bernoulli variance from lecture.

b. Find $\V[T]$ using the computational identity
   $\V[X] = \E[X^2] - (\E[X])^2$. Explain why $\E[T^2] = \E[T]$ for any
   indicator $T$.

c. Show that $\Cov(C, T) = \E[CT] - \E[C]\,\E[T]$, following the steps
   lecture used for the computational identity. Then compute $\Cov(C, T)$.
   Campaigns choose whom to contact. Name one way they choose that would
   produce this covariance even if contact had no effect on turnout.

d. Let $K = C + T$. Find $\V[K]$ using the variance of a sum. Then find the
   distribution of $K$ directly and confirm your answer. Report the value of
   $\V[K]$ you would get by wrongly treating $C$ and $T$ as independent.

e. A survey records a respondent's self-placement as $X = -1$ (left), $0$
   (centre) or $1$ (right), each with probability $1/3$. Let $Y = X^2$. $Y$
   equals 1 if the respondent takes a side and 0 if they place themselves at
   the centre. Show that $\Cov(X, Y) = 0$. Then show that $X$ and $Y$ are not
   independent by finding values $x$ and $y$ with
   $\Prob(X = x, Y = y) \neq \Prob(X = x)\,\Prob(Y = y)$.
