---
title: "{{< var course.code >}}: Homework 3 — Probability"
subtitle: "{{< var course.semester >}}"
author:
  - "{{< var course.instructor >}} (Instructor)"
  - "{{< var course.ta >}} (Teaching Assistant)"
format:
  pdf:
    documentclass: article
    pdf-engine: pdflatex
    fontsize: 11pt
    geometry: margin=1in
    colorlinks: true
    include-in-header:
      - file: ../../theme/macros.tex
execute:
  echo: true
  eval: false
---

# Part A — The rules of probability

## Question 1: Two elections

One registered voter is drawn at random from a county's voter file. Let $A$ be
the event that they voted in the spring primary and $B$ the event that they
voted in the November general election. In this county,

$$
\Prob(A) = .24, \qquad \Prob(B) = .62, \qquad \Prob(A \cap B) = .21 .
$$

a. State the sample space for this draw, and say what $A^c$ and $A \cap B$
   each describe in words.

b. Compute $\Prob(A \cup B)$.

c. Compute the probability that the voter cast a ballot in neither election.

d. Compute the probability that the voter cast a ballot in **exactly one** of
   the two. Get to the answer twice: once from your answer to (b), and once by
   computing $\Prob(A \cap B^c)$ and $\Prob(A^c \cap B)$ separately.

e. A colleague building the same table reports $\Prob(A \cap B) = .30$
   instead. Explain what the problem with that number is.

## Question 2: Regimes and leader exits

Country-years between 1946 and 2020 are cross-classified by regime type and by
whether the sitting leader left office through irregular means — a coup, a
forced resignation, an assassination. Treat the table as the joint
distribution for one country-year drawn at random. (These proportions are
stipulated for the exercise.)

\begin{center}
\begin{tabular}{@{}lrrr@{}}
\toprule
 & Irregular exit & No irregular exit & \textbf{Total} \\
\midrule
Democracy & .002 & .548 & \textbf{.550} \\
Autocracy & .018 & .432 & \textbf{.450} \\
\midrule
\textbf{Total} & \textbf{.020} & \textbf{.980} & \textbf{1.000} \\
\bottomrule
\end{tabular}
\end{center}

a. Report the two marginal distributions, and say which cells of the table you
   summed to get each one.

b. Compute $\Prob(\text{irregular} \mid \text{autocracy})$ and
   $\Prob(\text{irregular} \mid \text{democracy})$.

c. Compute $\Prob(\text{autocracy} \mid \text{irregular})$.

d. Your answers to (b) and (c) are both about autocracy and irregular exits,
   and they are far apart. Explain in two or three sentences what each one
   measures, and which of the two answers the question "how dangerous is it to
   be an autocrat?"

e. Determine whether regime type and irregular exit are independent. Show the
   comparison the definition requires.

## Question 3: Two kinds of sampling

**(i) An audit.** A county has 40 precincts, 6 of which contain a
discrepancy between the machine count and the paper record. An audit selects 3
precincts at random **without replacement**.

a. Compute the probability that the first precinct selected contains a
   discrepancy.

b. Compute the probability that the first two both contain discrepancies.

c. Compute the probability that **none** of the 3 selected precincts contains
   a discrepancy, and then the probability that at least one does.

d. Recompute (c) on the assumption that the three draws are made **with**
   replacement. Report both numbers to three decimal places, say which
   procedure is more likely to catch at least one discrepancy, and explain in
   one sentence why the two answers differ.

**(ii) Replications.** An early field experiment reports that door-to-door
canvassing raises turnout. Five research teams set out to check the result.
Each runs its own canvassing experiment in a different city, drawing its own
random sample of households and running its own randomization.

For this question, assume the original finding is correct: canvassing does
raise turnout. Assume also that each team's experiment has an 80 percent
chance of reproducing it, and that the five experiments are independent of one
another. Call a team that reproduces the finding a **success**.

e. Compute the probability that all five teams succeed.

f. Compute the probability that at least one of the five does not.

g. Compute the probability that all $k$ teams succeed for
   $k = 5, 8, 10, 11, 12$. Then state the smallest $k$ for which that
   probability falls below .10.

h. The independence assumed here is a property of how these five experiments
   were run, not a claim about how politics works. Say where it comes from,
   then describe one concrete feature a set of replications could have that
   would break it.

# Part B — Bayes' rule

## Question 4: Signature verification

A county processes 200,000 mail ballots. One ballot in 10,000 carries a
forged signature. The county's signature-matching system flags a forged
signature with probability .95, and flags a legitimate signature as forged
with probability .02.

a. Let $F$ be the event that a ballot's signature is forged and $\text{Flag}$
   the event that the system flags it. Write the quantity the county cares
   about as a conditional probability, and write out Bayes' rule for it with
   the denominator expanded.

b. Compute $\Prob(F \mid \text{Flag})$.

c. Redo the calculation in counts rather than probabilities. Of the 200,000
   ballots, how many are forged, how many of those are flagged, how many
   legitimate ballots are flagged, and how many ballots are flagged in total?
   Confirm that the ratio of the relevant two counts reproduces your answer to
   (b).

d. The county tightens the system so that it flags a legitimate signature as
   forged with probability .002, leaving everything else unchanged. Recompute
   $\Prob(F \mid \text{Flag})$.

e. Every flagged ballot is set aside until the voter is contacted and
   confirms it. Using your counts from (c), write two or three sentences on
   what the base rate implies for that process.

# Part C — Random variables and distributions

## Question 5: Three coin flips, one experiment

A survey experiment has three subjects. Each subject is assigned to treatment
or control by an independent flip of a fair coin. Let $X$ be the number of
subjects assigned to treatment.

If a part below has no appropriate answer, say so and explain why.

a. List the sample space $\Omega$ of assignment outcomes, and give the
   probability of each outcome.

b. Give the value of $X$ for each outcome in $\Omega$, then write the
   probability mass function $f(x) = \Prob(X = x)$ as a table.

c. Confirm that your answer to (b) satisfies all three conditions in the
   definition of a PMF.

d. Write the cumulative distribution function $F(x) = \Prob(X \le x)$ at
   $x = 0, 1, 2, 3$. Then give $F(1.5)$ and $F(-2)$.

e. Give the probability density function of $X$.

f. Name the distribution $X$ follows, with its parameters. (Hint: see slide
   44 in the Week 3 lecture slides.)

g. Compute the probability that the experiment assigns all three subjects to
   the same condition. Then say, in one sentence, what that number implies
   about assigning treatment by independent coin flips in an experiment this
   small.

# Part D — Applied: education and party in the VOTER data

```{r}
library(tidyverse)
source("scripts/gov6019_ggplot.R")        # adjust to your project's layout
voter <- read_csv("data/voter2012.csv")   # adjust to your project's layout
```

## Question 6: A joint distribution estimated from data

Collapse `educ` into a binary variable separating respondents who completed a
four-year degree or more from everyone else, then cross-classify by that
variable and `pid3`. Drop rows missing either variable, and label both factors
with words.

a. Consult the codebook and say which values of `educ` you put on each side of
   the cut. How many rows does the complete-case table contain, and how many
   are dropped? Report how many of the dropped rows are missing `educ`, how
   many are missing `pid3`, and how many are missing both. State the
   assumption you make about those respondents by dropping them.

b. Report the joint distribution as a 2 $\times$ 3 table of relative
   frequencies over the complete cases, with row and column totals. Confirm
   that the six cells sum to 1.

c. Read $\Prob(\text{Democrat})$ and $\Prob(\text{BA or higher})$ off your
   table, and say which cells you summed for each.

d. Compute $\Prob(\text{Democrat} \mid \text{BA or higher})$ and
   $\Prob(\text{Democrat} \mid \text{No BA})$ from the table. Then compute
   $\Prob(\text{BA or higher} \mid \text{Democrat})$.

e. Assess whether party identification and education are independent in this
   sample. Compare the six joint relative frequencies against the six products
   of the corresponding marginals, and report the largest discrepancy. Say
   which cells are over-represented relative to independence and which are
   under-represented.

f. Your answer to (e) is a statement about 7,895 people who answered a survey
   in 2012. Say what would have to be true for it to be a statement about the
   American electorate, and name one quantity you would need in order to
   decide whether a discrepancy of the size you found in (e) is more than an
   artifact of which 8,000 people happened to be sampled.

g. Make one figure showing the conditional distribution of party
   identification given education. Label the factor levels with words, and
   start any bars at zero. Add a two-sentence caption: the first sentence
   states what the figure shows, the second states a limitation of the data
   behind it.

   The shares you plot are the ones from (d), not raw counts: the two
   education groups differ in size, so counts are not comparable across them.

   ```{r}
   # within-group shares: count the cells, then divide by the group's own total
   shares <- cc |>
     count(ba_f, pid3_f) |>
     group_by(ba_f) |>
     mutate(share = n / sum(n)) |>
     ungroup()

   # side-by-side bars rather than stacked:
   #   geom_col(position = position_dodge(width = 0.7), width = 0.6)
   # percent labels on the y axis:
   #   scale_y_continuous(labels = scales::label_percent(accuracy = 1))
   # a title that states the finding, and no axis left unlabeled:
   #   labs(title = "...", x = NULL, y = "...")
   ```

---

*If you want more:* Blitzstein and Hwang, *Introduction to Probability*
(free at <https://probabilitybook.net/>), chapters 1--3, covers Parts A--C
at and somewhat beyond this level; Gailmard, ch. 3, is the assigned
treatment; Tversky and Kahneman, "Judgment under Uncertainty: Heuristics and
Biases," *Science* 185(4157), 1974, is the origin of the base-rate problem
Question 4 sets up; and the Open Science Collaboration, "Estimating the
Reproducibility of Psychological Science," *Science* 349(6251), 2015, is the
study Question 3(ii) is modeled on.
